---
title: "Hands on Exercise 4B : Visual Statistical Analysis"
author: "Gautamgovan"
date: "February 5, 2026"
format: 
  html:
    toc: true
    toc-depth: 4
    number-sections: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview

In this hands-on exercise, we will gain hands-on experience on using:

-   ggstatsplot package to create visual graphics with rich statistical information,

-   performance package to visualise model diagnostics, and

-   parameters package to visualise model parameters

## Visual Statistical Analysis with ggstatsplot

ggstatsplot is an extension of ggplot2 package for creating graphics with details from statistical tests included in the information-rich plots themselves.

-   To provide alternative statistical inference methods by default.
-   To follow best practices for statistical reporting. For all statistical tests

## Getting Started

### Installing and launching R packages

In this exercise, ggstatsplot and tidyverse will be used.

```{r}
pacman::p_load(ggstatsplot, tidyverse)
```

### Importing data

We import Exam.csv data

```{r}
exam <- read_csv("data/Exam_data.csv")
```

### One-sample test: gghistostats() method

In the code chunk below, gghistostats() is used to to build an visual of one-sample test on English scores.

```{r}
set.seed(1234)

gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores"
)
```

Default information: - statistical details - Bayes Factor - sample sizes - distribution summary

### Unpacking the Bayes Factor

-   A Bayes factor is the ratio of the likelihood of one particular hypothesis to the likelihood of another. It can be interpreted as a measure of the strength of evidence in favor of one theory among two competing theories.

-   That’s because the Bayes factor gives us a way to evaluate the data in favor of a null hypothesis, and to use external information to do so. It tells us what the weight of the evidence is in favor of a given hypothesis.

-   When we are comparing two hypotheses, ( H_1 ) (the alternate hypothesis) and ( H_0 ) (the null hypothesis), the Bayes Factor is often written as ( B\_{10} ). It can be defined mathematically as

\[ \frac{\text{likelihood of data given } H_1} {\text{likelihood of data given } H_0} = \frac{P(D \mid H_1)}{P(D \mid H_0)} \]

### How to interpret Bayes Factor

A Bayes Factor can be any positive number. One of the most common interpretations is this one—first proposed by Harold Jeffereys (1961) and slightly modified by Lee and Wagenmakers in 2013:

| If (B\_{10}) is… | Then you have…                 |
|------------------|--------------------------------|
| (\> 100)         | Extreme evidence for (H_1)     |
| (30 - 100)       | Very strong evidence for (H_1) |
| (10 - 30)        | Strong evidence for (H_1)      |
| (3 - 10)         | Moderate evidence for (H_1)    |
| (1 - 3)          | Anecdotal evidence for (H_1)   |
| \(1\)            | No evidence                    |
| (1/3 - 1)        | Anecdotal evidence for (H_0)   |
| (1/10 - 1/3)     | Moderate evidence for (H_0)    |
| (1/30 - 1/10)    | Strong evidence for (H_0)      |
| (1/100 - 1/30)   | Very strong evidence for (H_0) |
| (\< 1/100)       | Extreme evidence for (H_0)     |

### Two-sample mean test: ggbetweenstats()

In the code chunk below, ggbetweenstats() is used to build a visual for two-sample mean test of Maths scores by gender.

```{r}
ggbetweenstats(
  data = exam,
  x = GENDER, 
  y = MATHS,
  type = "np",
  messages = FALSE
)
```

### Oneway ANOVA Test: ggbetweenstats() method

In the code chunk below, ggbetweenstats() is used to build a visual for One-way ANOVA test on English score by race.

```{r}
ggbetweenstats(
  data = exam,
  x = RACE, 
  y = ENGLISH,
  type = "p",
  mean.ci = TRUE, 
  pairwise.comparisons = TRUE, 
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```

-   “ns” → only non-significant
-   “s” → only significant
-   “all” → everything

#### ggbetweenstats - Summary of tests

Between-Subjects Tests

The following (between-subjects) tests are carried out for each type of analysis.

| Type | No. of groups | Test |
|----|----|----|
| Parametric | \> 2 | Fisher’s or Welch’s one-way ANOVA |
| Non-parametric | \> 2 | Kruskal–Wallis one-way ANOVA |
| Robust | \> 2 | Heteroscedastic one-way ANOVA for trimmed means |
| Bayes Factor | \> 2 | Fisher’s ANOVA |
| Parametric | 2 | Student’s or Welch’s *t*-test |
| Non-parametric | 2 | Mann–Whitney *U* test |
| Robust | 2 | Yuen’s test for trimmed means |
| Bayes Factor | 2 | Student’s *t*-test |

------------------------------------------------------------------------

Effect Sizes and Confidence Intervals

The following effect sizes (and confidence intervals, CI) are available for each type of test.

| Type | No. of groups | Effect size | CI? |
|----------------|-----------------------|-----------------|----------------|
| Parametric | \> 2 | (\eta\_p\^2), (\eta\^2), (\omega\_p\^2), (\omega\^2) | Yes |
| Non-parametric | \> 2 | (\eta\_H\^2) (H-statistic–based eta-squared) | Yes |
| Robust | \> 2 | (\xi) (explanatory measure of effect size) | Yes |
| Bayes Factor | \> 2 | No | No |
| Parametric | 2 | Cohen’s *d*, Hedges’ *g* (central and noncentral *t*-based) | Yes |
| Non-parametric | 2 | \(r\) (computed as (Z / \sqrt{N})) | Yes |
| Robust | 2 | (\xi) (explanatory measure of effect size) | Yes |
| Bayes Factor | 2 | No | No |

------------------------------------------------------------------------

Multiple Pairwise Comparison Tests

The following multiple pairwise comparison tests are supported.

| Type | Equal variance? | Test | *p*-value adjustment? |
|----|----|----|----|
| Parametric | No | Games–Howell test | Yes |
| Parametric | Yes | Student’s *t*-test | Yes |
| Non-parametric | No | Dunn test | Yes |
| Robust | No | Yuen’s trimmed means test | Yes |
| Bayes Factor | NA | Student’s *t*-test | NA |

### Significant Test of Correlation: ggscatterstats()

In the code chunk below, ggscatterstats() is used to build a visual for Significant Test of Correlation between Maths scores and English scores.

```{r}

ggscatterstats(
  data = exam,
  x = MATHS,
  y = ENGLISH,
  marginal = FALSE,
  )
```

### Significant Test of Association (Depedence) : ggbarstats() methods

In the code chunk below, the Maths scores is binned into a 4-class variable by using cut().

```{r}
exam1 <- exam %>% 
  mutate(MATHS_bins = 
           cut(MATHS, 
               breaks = c(0,60,75,85,100))
)
```

In this code chunk below ggbarstats() is used to build a visual for Significant Test of Association

```{r}
ggbarstats(exam1, 
           x = MATHS_bins, 
           y = GENDER)
```
