---
title: "Take Home Exercise 2B"
author: "Gautamgovan"
date: "March 4, 2026"
date-modified: "last-modified"
format: 
  html:
    toc: true
    toc-depth: 4
    number-sections: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Objective:
The objective of this project is to understand the behaviors behind customer transactions and how socio-demographic factors—such as age, income, and occupation—affect customer transactions. By applying statistical correlation techniques, we aim to identify which personal attributes are the true drivers of financial activity within the Colombian fintech ecosystem.

This analysis extends the exploratory study by developing predictive models to estimate customer financial activity within the Colombian fintech ecosystem. Instead of focusing on customer satisfaction, this stage models transaction behaviour, which more directly reflects financial engagement on the platform. Specifically, the study aims to predict the number of average transactions per day performed by each customer using demographic attributes, digital engagement indicators, and behavioural transaction features. By applying machine learning models such as Linear Regression, Decision Tree, and Random Forest, the analysis seeks to identify the key drivers of financial activity and evaluate the predictive capability of customer attributes in explaining transaction behaviour.

## Load Packages 

Several R packages are used in this analysis to support data manipulation, visualization, and predictive modelling. The tidyverse package is used for data cleaning and transformation, while lubridate assists with handling date and time variables in the transaction dataset. The tidymodels framework provides a unified environment for building and evaluating predictive models. Visualization packages such as corrplot, vip, and rpart.plot are used to produce visual analytics outputs, including correlation matrices, feature importance plots, and decision tree diagrams. These packages collectively support the implementation of the modelling workflow and the generation of interpretable visual insights.

```{r}
pacman::p_load(
  tidyverse,
  tidymodels,
  lubridate,
  vip,
  rpart.plot,
  corrplot,
  patchwork,
  scales,
  caret,
  doParallel
)

tidymodels_prefer()
theme_set(theme_minimal())
set.seed(1234)

```
## Loading Dataset

The **COFINFAD (Colombian Fintech Financial Analytics Dataset)** consists of two relational datasets provided in CSV format.

The first dataset, `customer_data.csv`, contains customer-level information for 48,723 individuals, including demographic attributes, product usage indicators, satisfaction metrics, and aggregated behavioral variables.

The second dataset, `transactions_data.csv`, contains 3,159,157 transaction records collected over a 12-month period in 2023, capturing individual transaction events such as transaction amount, transaction date, and transaction type.

Both datasets are linked through a common identifier, `customer_id`, which serves as the primary key in the customer dataset and as a foreign key in the transaction dataset.

In this analysis, the datasets are loaded into the  environment using the `readr` package, after which transaction-level data can be aggregated and merged with the customer-level dataset for further modelling and analysis.

```{r}
customer_data <- read_csv("data/customer_data.csv")

```
## Data Preparation
The first step is to remove observations with missing target values and ensure the transaction date field is in a proper date format.

```{r}
customer_data_clean <- customer_data %>%
  filter(!is.na(avg_daily_transactions))
```


### Inspect Dataset Structure 
```{r}
glimpse(customer_data_clean)
```
### Target Variable Distribution
Before building predictive models, it is important to examine the distribution of the target variable.
```{r}
customer_data_clean %>%
  summarise(
    min = min(avg_daily_transactions),
    mean = mean(avg_daily_transactions),
    median = median(avg_daily_transactions),
    max = max(avg_daily_transactions),
    sd = sd(avg_daily_transactions)
  )
```


```{r}
# Create modelling dataset with log-transformed target
customer_model_data <- customer_data_clean %>%
  mutate(
    log_avg_daily_transactions = log1p(avg_daily_transactions)
  )
```

### Distribution of Original and Log-Transformed Target

To confirm whether the transformation improves the distributional shape, we compare the original and transformed target using histograms and summary statistics.

```{r}
# Summary statistics for original target
customer_model_data %>%
  summarise(
    min_target    = min(avg_daily_transactions, na.rm = TRUE),
    mean_target   = mean(avg_daily_transactions, na.rm = TRUE),
    median_target = median(avg_daily_transactions, na.rm = TRUE),
    max_target    = max(avg_daily_transactions, na.rm = TRUE),
    sd_target     = sd(avg_daily_transactions, na.rm = TRUE)
  )
```
```{r}
# Summary statistics for log-transformed target
customer_model_data %>%
  summarise(
    min_log    = min(log_avg_daily_transactions, na.rm = TRUE),
    mean_log   = mean(log_avg_daily_transactions, na.rm = TRUE),
    median_log = median(log_avg_daily_transactions, na.rm = TRUE),
    max_log    = max(log_avg_daily_transactions, na.rm = TRUE),
    sd_log     = sd(log_avg_daily_transactions, na.rm = TRUE)
  )
```
```{r}
# Histogram: original target
p1 <- ggplot(customer_model_data, aes(x = avg_daily_transactions)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Distribution of Avg Daily Transactions",
    x = "avg_daily_transactions",
    y = "Count"
  ) +
  theme_minimal()

# Histogram: log-transformed target
p2 <- ggplot(customer_model_data, aes(x = log_avg_daily_transactions)) +
  geom_histogram(bins = 40, fill = "darkorange", color = "white") +
  labs(
    title = "Distribution of Log-Transformed Avg Daily Transactions",
    x = "log1p(avg_daily_transactions)",
    y = "Count"
  ) +
  theme_minimal()

p1 + p2
```


The target variable avg_daily_transactions exhibited strong right skewness, with the majority of customers performing very few daily transactions while a small number of users generated extremely high values. Such skewness can negatively affect predictive modelling by allowing extreme observations to dominate model estimation. To mitigate this issue, a logarithmic transformation was applied using log1p(avg_daily_transactions), which compresses large values while preserving zero observations. The resulting distribution shows a substantially reduced skewness and a more stable range suitable for regression-based modelling.


## Feature Selection


Before building predictive models, it is necessary to select appropriate predictor variables. 
Certain variables in the dataset directly summarize transaction activity (such as transaction counts or transaction volumes). 
Including such variables would introduce **data leakage**, because they are mathematically related to the target variable `avg_daily_transactions`.

To avoid this issue, only demographic attributes, product ownership indicators, digital engagement metrics, and customer experience variables are retained as predictors.

```{r}
selected_vars <- c(
  "age",
  "gender",
  "income_bracket",
  "occupation",
  "education_level",
  "marital_status",
  "household_size",
  "customer_segment",
  "acquisition_channel",
  "customer_tenure",
  "savings_account",
  "credit_card",
  "personal_loan",
  "investment_account",
  "insurance_product",
  "active_products",
  "app_logins_frequency",
  "feature_usage_diversity",
  "bill_payment_user",
  "auto_savings_enabled",
  "credit_utilization_ratio",
  "international_transactions",
  "failed_transactions",
  "base_satisfaction",
  "product_satisfaction",
  "satisfaction_score",
  "nps_score",
  "support_tickets_count",
  "resolved_tickets_ratio",
  "app_store_rating",
  "feedback_sentiment",
  "feature_requests"
)

model_data <- customer_model_data %>%
  select(all_of(selected_vars), log_avg_daily_transactions)
```

## Data Splitting 

To ensure reliable evaluation, the dataset is divided into three subsets:

- **Training set (70%)** – used to train the models  
- **Validation set (15%)** – used for hyperparameter tuning  
- **Test set (15%)** – used for final model evaluation

```{r}
set.seed(1234)

data_split <- initial_split(model_data, prop = 0.70)

train_data <- training(data_split)
temp_data <- testing(data_split)

validation_split <- initial_split(temp_data, prop = 0.50)

validation_data <- training(validation_split)
test_data <- testing(validation_split)
```

```{r}
#| include: false
get_or_save_rds <- function(path, expr) {
  if (file.exists(path)) {
    readRDS(path)
  } else {
    obj <- eval.parent(substitute(expr))
    saveRDS(obj, path)
    obj
  }
}
```


## Preprocessing Recipe

A preprocessing recipe is created using the **tidymodels recipe framework**. 
This step prepares the dataset for modelling by encoding categorical variables and removing predictors with near-zero variance.

```{r}
rec <- recipe(log_avg_daily_transactions ~ ., data = train_data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

```

## Linear Regression Model 

Linear regression is used as a baseline model to estimate the relationship between customer attributes and transaction activity.

```{r}
lm_model <- linear_reg() %>%
  set_engine("lm")

lm_workflow <- workflow() %>%
  add_model(lm_model) %>%
  add_recipe(rec)

lm_fit <- get_or_save_rds(
  "models/lm_fit.rds",
  fit(lm_workflow, data = train_data)
)

```
## Decision Tree Model

Decision trees are able to capture non-linear relationships between predictors and transaction behaviour.


```{r}
tree_model <- decision_tree(
  tree_depth = tune(),
  cost_complexity = tune(),
  min_n = tune()
) %>%
  set_engine("rpart") %>%
  set_mode("regression")
```

```{r}
tree_grid <- grid_regular(
  tree_depth(range = c(2L, 8L)),
  cost_complexity(range = c(-4, -2)),
  min_n(range = c(5L, 20L)),
  levels = 3
)


```

```{r}
tree_workflow <- workflow() %>%
  add_recipe(rec) %>%
  add_model(tree_model)
cv_folds <- vfold_cv(train_data, v = 5)

```

### Tune Decision Tree on Validation Set
```{r}
tree_model_path <- "models/tree_tuned.rds"

if (file.exists(tree_model_path)) {

  tree_tuned <- readRDS(tree_model_path)

} else {

  tree_tuned <- tune_grid(
  tree_workflow,
  resamples = cv_folds,
  grid = tree_grid,
  metrics = metric_set(rmse, mae, rsq),
  control = control_grid(verbose = TRUE)
)


  saveRDS(tree_tuned, tree_model_path)

}

```

### Best Tree 
```{r}
best_tree <- select_best(tree_tuned, metric = "rmse")

best_tree
```
### Final Tree Model 

```{r}
tree_final <- finalize_workflow(tree_workflow, best_tree) %>%
  fit(train_data)
```

### Plot Tree 
```{r}
rpart.plot(extract_fit_engine(tree_final))
```

## Random Forest

### Register Parallel Processing

```{r}
library(parallel)
library(doParallel)

n_cores <- max(1, detectCores() - 1)
cl <- makePSOCKcluster(n_cores)
registerDoParallel(cl)

n_cores
```
### Prepare Predictor Count for mtry

Because dummy variables expand the number of predictors, the recipe is preprocessed once to estimate the usable predictor count.

```{r}
rec_prep <- prep(rec)
train_processed <- bake(rec_prep, new_data = train_data)

num_predictors <- ncol(train_processed) - 1
num_predictors
```
### Random Forest Specification

Use fewer trees during tuning for speed, then a larger number for the final fitted model.

```{r}
rf_model <- rand_forest(
  trees = 200,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", importance = "impurity", num.threads = n_cores) %>%
  set_mode("regression")
```

### Random Forest Workflow

```{r}
rf_workflow <- workflow() %>%
  add_recipe(rec) %>%
  add_model(rf_model)
```


### Random Forest Grid


```{r}
rf_grid <- crossing(
  mtry = unique(pmax(1, round(c(
    sqrt(num_predictors),
    num_predictors / 3,
    num_predictors / 2
  )))),
  min_n = c(5, 10, 20)
)

rf_grid
```
### Tune Random Forest

```{r}
rf_model_path <- "models/rf_tuned.rds"

if (file.exists(rf_model_path)) {

  rf_tuned <- readRDS(rf_model_path)

} else {

  rf_tuned <- tune_grid(
    rf_workflow,
    resamples = cv_folds,
    grid = rf_grid,
    metrics = metric_set(rmse, mae, rsq),
    control = control_grid(
      verbose = TRUE,
      save_pred = FALSE,
      allow_par = TRUE
    )
  )

  saveRDS(rf_tuned, rf_model_path)

}
```

### Best Random Forest Model

```{r}
best_rf <- select_best(rf_tuned, metric = "rmse")

best_rf
```
### Final Random Forest Model

Use more trees for the final model.

```{r}
rf_final_path <- "models/rf_final.rds"

if (file.exists(rf_final_path)) {

  rf_final <- readRDS(rf_final_path)

} else {

  rf_final_model <- rand_forest(
    trees = 400,
    mtry = best_rf$mtry,
    min_n = best_rf$min_n
  ) %>%
    set_engine("ranger", importance = "impurity", num.threads = n_cores) %>%
    set_mode("regression")

  rf_final_workflow <- workflow() %>%
    add_recipe(rec) %>%
    add_model(rf_final_model)

  rf_final <- fit(rf_final_workflow, data = train_data)

  saveRDS(rf_final, rf_final_path)

}
```

## Model Performance Comparison

The three models are evaluated on the validation set first so that their predictive performance can be compared consistently.

### Generate Validation Predictions

```{r}
lm_val_preds <- predict(lm_fit, new_data = validation_data) %>%
  bind_cols(validation_data %>% select(log_avg_daily_transactions)) %>%
  mutate(model = "Linear Regression")

tree_val_preds <- predict(tree_final, new_data = validation_data) %>%
  bind_cols(validation_data %>% select(log_avg_daily_transactions)) %>%
  mutate(model = "Decision Tree")

rf_val_preds <- predict(rf_final, new_data = validation_data) %>%
  bind_cols(validation_data %>% select(log_avg_daily_transactions)) %>%
  mutate(model = "Random Forest")
```
### Combine Predictions 

```{r}
all_val_preds <- bind_rows(
  lm_val_preds,
  tree_val_preds,
  rf_val_preds
)
```


### Compute Metrics 

```{r}
model_metrics <- all_val_preds %>%
  group_by(model) %>%
  summarise(
    RMSE = rmse_vec(log_avg_daily_transactions, .pred),
    MAE  = mae_vec(log_avg_daily_transactions, .pred),
    R2   = rsq_vec(log_avg_daily_transactions, .pred)
  )

model_metrics
```
### Comparsion Chat 

```{r}
model_metrics_long <- model_metrics %>%
  mutate(
    model = factor(
      model,
      levels = c("Decision Tree", "Linear Regression", "Random Forest")
    )
  ) %>%
  pivot_longer(
    cols = c(RMSE, MAE, R2),
    names_to = "Metric",
    values_to = "Value"
  )

ggplot(model_metrics_long, aes(x = model, y = Value, fill = model)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  facet_wrap(~ Metric, scales = "free_y", nrow = 1) +
  scale_fill_brewer(palette = "Set2") +
  scale_x_discrete(labels = c("Decision\nTree", "Linear\nRegression", "Random\nForest")) +
  labs(
    title = "Model Performance Comparison on Validation Set",
    x = "Model",
    y = "Metric Value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(size = 11, angle = 0, vjust = 1),
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 12, face = "bold"),
    plot.margin = margin(10, 10, 20, 10)
  )


```

### Visual Analytics for the Best Model

#### Feature Importance for Random Forest

```{r}
vip(extract_fit_parsnip(rf_final)$fit, num_features = 15)
```

#### Predicted vs Actual Values for Random Forest

```{r}
ggplot(rf_val_preds, aes(x = log_avg_daily_transactions, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(
    title = "Predicted vs Actual Values for Random Forest",
    x = "Actual Log Avg Daily Transactions",
    y = "Predicted Log Avg Daily Transactions"
  )
```

#### Residual Plot for Random Forest

```{r}
rf_val_preds <- rf_val_preds %>%
  mutate(residual = log_avg_daily_transactions - .pred)

ggplot(rf_val_preds, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot for Random Forest",
    x = "Predicted Values",
    y = "Residuals"
  )
```



## Final Test Set Evaluation

After model comparison on the validation set, the best-performing model can be assessed on the unseen test set.

### Test Predictions for Best Model

```{r}

rf_test_preds <- predict(rf_final, new_data = test_data) %>%
  bind_cols(test_data %>% select(log_avg_daily_transactions))
```

### Test Metrics 

```{r}
rf_test_metrics <- tibble(
  RMSE = rmse_vec(rf_test_preds$log_avg_daily_transactions, rf_test_preds$.pred),
  MAE  = mae_vec(rf_test_preds$log_avg_daily_transactions, rf_test_preds$.pred),
  R2   = rsq_vec(rf_test_preds$log_avg_daily_transactions, rf_test_preds$.pred)
)

rf_test_metrics
```

```{r}
#| include: false
stopCluster(cl)
registerDoSEQ()
```

## Conclusion from Predicting Customer Transaction Behaviour


This analysis aimed to identify the drivers of financial activity in the Colombian fintech ecosystem by predicting customer transaction behaviour, measured using average daily transactions. Three models—Linear Regression, Decision Tree, and Random Forest were developed using demographic, product usage, and digital engagement variables from the COFINFAD dataset.

Among the models, Random Forest achieved the best overall predictive performance, with slightly lower RMSE and MAE and the highest R² value compared with the other models. This suggests that ensemble methods are better able to capture complex relationships between customer attributes and transaction behaviour.

However, the overall explanatory power of the models remained moderate, indicating that the available predictors explain only part of the variation in customer transaction activity. This suggests that additional behavioural or contextual factors may influence customer financial activity.

Overall, the results demonstrate that machine learning models can provide useful insights into transaction behaviour, while highlighting opportunities to improve predictive performance with richer behavioural data.

