---
title: "Hands on Exercise 5D:Visual Multivariate Analysis with Parallel Coordinates Plot"
author: "Gautamgovan"
date: "February 12, 2026"
date-modified: "last-modified"
format: 
  html:
    toc: true
    toc-depth: 4
    number-sections: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview

Parallel coordinates plot is a data visualisation specially designed for visualising and analysing multivariate, numerical data.

By the end of this hands-on exercise, we will gain hands-on experience on:

- plotting statistic parallel coordinates plots by using ggparcoord() of GGally package, and
- plotting interactive parallel coordinates plots by using parallelPlot package.

## Installing and Launching R Packages

For this exercise, the GGally, parallelPlot and tidyverse packages will be used.

The code chunks below are used to install and load the packages in R.

```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```
##  Data Preparation
In this hands-on exercise, the World Happinees 2018  data will be used.

In the code chunk below, read_csv() of readr package is used to import WHData-2018.csv into R and save it into a tibble data frame object called wh.

```{r}
wh <- read_csv("data/WHData-2018.csv")
```
## Plotting Static Parallel Coordinates Plot

In this section, you will learn how to plot static parallel coordinates plot by using ggparcoord() of GGally package. Before getting started, it is a good practice to read the function description in detail.

### Plotting a simple parallel coordinates

Code chunk below shows a typical syntax used to plot a basic static parallel coordinates plot by using ggparcoord().

```{r}
ggparcoord(data = wh, 
           columns = c(7:12))
```
Notice that only two argument namely data and columns is used. Data argument is used to map the data object (i.e. wh) and columns is used to select the columns for preparing the parallel coordinates plot.

### Plotting a parallel coordinates with boxplot

The basic parallel coordinates failed to reveal any meaning understanding of the World Happiness measures. In this section, you will learn how to makeover the plot by using a collection of arguments provided by ggparcoord().

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")
```
### Parallel coordinates with facet
Since ggparcoord() is developed by extending ggplot2 package, we can combination use some of the ggplot2 function when plotting a parallel coordinates plot.

In the code chunk below, facet_wrap() of ggplot2 is used to plot 10 small multiple parallel coordinates plots. Each plot represent one geographical region such as East Asia.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)
```
### Rotating x-axis text label

To make the x-axis text label easy to read, let us rotate the labels by 30 degrees. We can rotate axis text labels using theme() function in ggplot2 as shown in the code chunk below

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30))
```

### Adjusting the rotated x-axis text label

Rotating x-axis text labels to 30 degrees makes the label overlap with the plot and we can avoid this by adjusting the text location using hjust argument to theme’s text element with element_text(). We use axis.text.x as we want to change the look of x-axis text.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```

## Plotting Interactive Parallel Coordinates Plot: parallelPlot methods

parallelPlot is an R package specially designed to plot a parallel coordinates plot by using ‘htmlwidgets’ package and d3.js. In this section, you will learn how to use functions provided in parallelPlot package to build interactive parallel coordinates plot.

### The basic plot
he code chunk below plot an interactive parallel coordinates plot by using parallelPlot().

```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))
parallelPlot(wh,
             width = 320,
             height = 250)
```
### Rotate axis label

In the code chunk below, rotateTitle argument is used to avoid overlapping axis labels.

```{r}
parallelPlot(wh,
             rotateTitle = TRUE)
```
One of the useful interactive feature of parallelPlot is we can click on a variable of interest, for example Happiness score, the monotonous blue colour (default) will change a blues with different intensity colour scheme will be used.

### Changing the colour scheme
We can change the default blue colour scheme by using continousCS argument as shown in the code chunk below.

```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)
```
### Parallel coordinates plot with histogram
In the code chunk below, histoVisibility argument is used to plot histogram along the axis of each variables.

```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)
```











